---
title: '20-1217_Conway_Cubes'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

#disable scientific notation
options(scipen = 999)
```

## Day 17: Conway Cubes

As your flight slowly drifts through the sky, the Elves at the Mythical Information Bureau at the North Pole contact you. They'd like some help debugging a malfunctioning experimental energy source aboard one of their super-secret imaging satellites.

The experimental energy source is based on cutting-edge technology: a set of Conway Cubes contained in a pocket dimension! When you hear it's having problems, you can't help but agree to take a look.

The pocket dimension contains an infinite 3-dimensional grid. At every integer 3-dimensional coordinate `(x,y,z)`, there exists a single cube which is either *active* or *inactive*.

In the initial state of the pocket dimension, almost all cubes start *inactive*. The only exception to this is a small flat region of cubes (your puzzle input); the cubes in this region start in the specified *active (#)* or *inactive (.)* state.

The energy source then proceeds to boot up by executing *six cycles*.

Each cube only ever considers its `neighbors`: any of the 26 other cubes where any of their coordinates differ by at most 1. For example, given the cube at x=1,y=2,z=3, its neighbors include the cube at x=2,y=2,z=2, the cube at x=0,y=2,z=3, and so on.

During a cycle, *all cubes simultaneously change their state* according to the following rules:

- If a cube is `active` and `exactly 2 or 3` of its neighbors are also active, the cube remains `active`. Otherwise, the cube becomes `inactive`.
- If a cube is `inactive` but `exactly 3` of its neighbors are `active`, the cube becomes `active`. Otherwise, the cube remains `inactive`.

The engineers responsible for this experimental energy source would like you to simulate the pocket dimension and determine what the configuration of cubes should be at the end of the six-cycle boot process.

For example, consider the following initial state:

`.#.`
`..#`
`###`

Even though the pocket dimension is 3-dimensional, this initial state represents a small 2-dimensional slice of it. (In particular, this initial state defines a 3x3x1 region of the 3-dimensional space.)

Simulating a few cycles from this initial state produces the following configurations, where the result of each cycle is shown layer-by-layer at each given z coordinate (and the frame of view follows the active cells in each cycle):

Before any cycles:

`z=0`
`.#.`
`..#`
`###`


After 1 cycle:

`z=-1`
`#..`
`..#`
`.#.`

`z=0`
`#.#`
`.##`
`.#.`

`z=1`
`#..`
`..#`
`.#.`


After 2 cycles:

`z=-2`
`.....`
`.....`
`..#..`
`.....`
`.....`

`z=-1`
`..#..`
`.#..#`
`....#`
`.#...`
`.....`

`z=0`
`##...`
`##...`
`#....`
`....#`
`.###.`

`z=1`
`..#..`
`.#..#`
`....#`
`.#...`
`.....`

`z=2`
`.....`
`.....`
`..#..`
`.....`
`.....`


After 3 cycles:

`z=-2`
`.......`
`.......`
`..##...`
`..###..`
`.......`
`.......`
`.......`

`z=-1`
`..#....`
`...#...`
`#......`
`.....##`
`.#...#.`
`..#.#..`
`...#...`

`z=0`
`...#...`
`.......`
`#......`
`.......`
`.....##`
`.##.#..`
`...#...`

`z=1`
`..#....`
`...#...`
`#......`
`.....##`
`.#...#.`
`..#.#..`
`...#...`

`z=2`
`.......`
`.......`
`..##...`
`..###..`
`.......`
`.......`
`.......`

After the full six-cycle boot process completes, `112` cubes are left in the active state.

Starting with your given initial configuration, simulate six cycles. 

*How many cubes are left in the active state after the sixth cycle?*

## INPUT

```{r input}
input_raw <- 
  read_lines("20-1217_input_conway_cubes.txt")

len <- str_length(input_raw[[1]])

input <- vector("list", len)

for (i in seq_along(input_raw)) {
  
  x_val <- input_raw[i] %>% 
    str_split(., "") %>% 
    .[[1]]
  
  
  input[[i]] <- 
    tibble(y = (len + 1) - i, z = 0, value = x_val) %>%
    mutate(x = row_number()) %>% 
    relocate(x)  
}

input <- 
  reduce(input, bind_rows) %>% 
  arrange(x, y, z)




```


## EXERCISE 1

```{r exercise 1, eval=FALSE}
# only doing six steps, so max z directions are -5 and 5

all_combos <- crossing(x = 1:len, y = 1:len, z = -6:6)

initial_state <- 
  left_join(all_combos, input, by = c("x", "y", "z")) %>% 
  mutate(value = replace_na(value, "."))

state <- initial_state

active_neighbors <- function(df, x1, y1, z1) { 
  df %>% 
    filter(between(x, max(x1 - 1, 1), min(x1 + 1, 8)), 
           between(y, max(y1 - 1, 1), min(y1 + 1, 8)), 
           between(z, max(z1 - 1, -6), min(z1 + 1, 6)), 
           str_c(x, y, z) != str_c(x1, y1, z1)) %>% 
    pull(value) %>%
    `==` (., "#") %>%
    sum()
}


conway_cycle <- function(.df, ...) {
  
  result <- .df
  
  len_cc <- length(unique(.df$x))
  
  for(i in 1:len_cc) { #x
    for (j in 1:len_cc) { #y
      for (k in -6:6) { #z
        spot <- .df$value[.df$x == i &
                            .df$y == j &
                            .df$z == k]
        
        nbors <- active_neighbors(.df, i, j, k)
        
        if (spot == "#" && nbors %in% c(2,3)) {
          next
        } else if (spot == "#") {
          result$value[result$x == i &
                         result$y == j &
                         result$z == k] <- "."
        } else if (spot == "." && nbors == 3) {
          result$value[result$x == i &
                         result$y == j &
                         result$z == k] <- "#"
        }
        
      }
      
    }
    
  }
  
  result
}

tictoc::tic("accumulate res1")

res1 <- reduce(1:6, conway_cycle, .init = initial_state)

tictoc::toc()



sum(res1$value == "#")


res1

```

With the above code, I get `121`, which is incorrect. 

(chunk above set to `eval=FALSE` as it's incorrect and I moved on to a correct approach)

Trying again, this time with an array

```{r exercise 1 array}

sample_data <- 
  matrix(str_split(".#...####", "")[[1]], nrow = 3, byrow = TRUE)

# need to pad by our steps...6 in all directions

input_matrix <- 
  str_split(input_raw , "") %>% reduce(., c) %>% 
  matrix(nrow = 8, byrow = TRUE)


data_array <- input_matrix

# array indices are [row, column, slice]

# 8y x 8x x 13z
max_y <- nrow(data_array) + (6 * 2)
max_x <- ncol(data_array) + (6 * 2)
max_z <- (2 * 6) + 1

# create initial array, first with all inactive
state_array <- rep(".", prod(max_x, max_y,  max_z))
dim(state_array) <- c(max_y, max_x, max_z)

# our initial position should be the middle z "slice" -- with 13 slices that's 7
state_array[7:(7 + (nrow(data_array) - 1)),
            7:(7 + (ncol(data_array) - 1)),
            ceiling(max_z / 2)] <- data_array

state_array[, , 7]

find_neighbors_array <- function(arr, y, x, z) {
  
  # if the spot is active already, subtract one from final sum so we don't include it
  subtract_spot <- if_else(arr[y, x, z] == "#", -1, 0)
  
  delta <- -1:1
  
  # find deltas
  delta_y <- y + delta
  delta_x <- x + delta
  delta_z <- z + delta
  
  # ensure deltas are within bounds
  delta_y <- delta_y[between(delta_y, 1, max_y)]
  delta_x <- delta_x[between(delta_x, 1, max_x)]
  delta_z <- delta_z[between(delta_z, 1, max_z)]
  
  # find neighbors and subtract 1 is spot counts as active 
  sum(arr[delta_y, delta_x, delta_z] == "#") + subtract_spot
  
}

conway_cycle_array <- function(.arr, ...) {
  
  dx <- dim(.arr)
  
  result <- .arr
  
  for (i in seq_len(dx[1])) { # y, rows
    for (j in seq_len(dx[2])){ # x, cols
      for (k in seq_len(dx[3])) { # z, slices
        
        spot <- .arr[i, j, k]
        
        nbors <- find_neighbors_array(.arr, i, j, k)
        
        if (spot == "#") {
          if (nbors %in% c(2,3)) {
            next
          } else {
            result[i, j, k] <- "."
          }
        } else if (spot == ".") {
          if (nbors == 3) {
            result[i, j, k] <- "#"
          } else {
            next
          }
        }
        
      } # end z slice loop
    } # end x col loop
  } # end y row loop
  
  return(result)
  
}

res1_array <- 
  reduce(1:6, conway_cycle_array, .init = state_array)

sum(res1_array == "#")


```

Forgot to take into account that the grid expands in the x and y plane as well as in the z plane.  Took that into account and got an answer of `252`


## EXERCISE 2 NARRATIVE

Same as before, but in four dimensions.

## EXERCISE 2

```{r exercise 2}

sample_data <- 
  matrix(str_split(".#...####", "")[[1]], nrow = 3, byrow = TRUE)
# need to pad by our steps...6 in all directions

input_matrix <- 
  str_split(input_raw , "") %>% reduce(., c) %>% 
  matrix(nrow = 8, byrow = TRUE)

# can use this to swap between sample and actual data
data_cube <- input_matrix

# array indices are [row, column, slice, hypercube]

# 8y x 8x x 13z x 13h
max_y <- nrow(data_cube) + (6 * 2)
max_x <- ncol(data_cube) + (6 * 2)
max_z <- (2 * 6) + 1
max_h <- (2 * 6) + 1

# create initial array, first with all inactive
state_hcube <- rep(".", prod(max_x, max_y,  max_z, max_h))
dim(state_hcube) <- c(max_y, max_x, max_z, max_h)

# our initial position should be the middle z "slice" and h "cube"
state_hcube[7:(7 + (nrow(data_cube) - 1)),
            7:(7 + (ncol(data_cube) - 1)),
            ceiling(max_z / 2), 
            ceiling(max_h / 2)] <- data_cube

# state_hcube[, , 8, 8]

find_neighbors_hcube <- function(arr, y, x, z, h) {
  
  # if the spot is active already, subtract one from final sum so we don't include it
  subtract_spot <- if_else(arr[y, x, z, h] == "#", -1, 0)
  
  delta <- -1:1
  
  # find deltas
  delta_y <- y + delta
  delta_x <- x + delta
  delta_z <- z + delta
  delta_h <- h + delta
  
  # ensure deltas are within bounds
  delta_y <- delta_y[between(delta_y, 1, max_y)]
  delta_x <- delta_x[between(delta_x, 1, max_x)]
  delta_z <- delta_z[between(delta_z, 1, max_z)]
  delta_h <- delta_h[between(delta_h, 1, max_h)]
  
  # find neighbors and subtract 1 is spot counts as active 
  sum(arr[delta_y, delta_x, delta_z, delta_h] == "#") + subtract_spot
  
}



conway_cycle_cube <- function(.arr, ...) {
  
  dx <- dim(.arr)
  
  result <- .arr
  
  for (i in seq_len(dx[1])) { # y, rows
    for (j in seq_len(dx[2])){ # x, cols
      for (k in seq_len(dx[3])) { # z, slices
        for (l in seq_len(dx[4])) { # h, hypercubes
          
          
          
          spot <- .arr[i, j, k, l]
          
          nbors <- find_neighbors_hcube(.arr, i, j, k, l)
          
          if (spot == "#") {
            if (nbors %in% c(2,3)) {
              next
            } else {
              result[i, j, k, l] <- "."
            }
          } else if (spot == ".") {
            if (nbors == 3) {
              result[i, j, k, l] <- "#"
            } else {
              next
            }
          }
        } # end of h hypercube loop
      } # end z slice loop
    } # end x col loop
  } # end y row loop
  
  return(result)
  
}

# conway_cycle_cube(state_hcube)

res2_cube <- 
  reduce(1:6, conway_cycle_cube, .init = state_hcube)

sum(res2_cube == "#")

```

