---
title: '20-1206 Custom Customs'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Day 6:Custom Customs

As your flight approaches the regional airport where you'll switch to a much larger plane, customs declaration forms are distributed to the passengers.

The form asks a series of 26 yes-or-no questions marked `a` through `z`. All you need to do is identify the questions for which **anyone in your group** answers "yes". Since your group is just you, this doesn't take very long.

However, the person sitting next to you seems to be experiencing a language barrier and asks if you can help. For each of the people in their group, you write down the questions for which they answer "yes", one per line. For example:

`abcx`
`abcy`
`abcz`

In this group, there are **6** questions to which anyone answered "yes": `a`, `b`, `c`, `x`, `y`, and `z`. (Duplicate answers to the same question don't count extra; each question counts at most once.)

Another group asks for your help, then another, and eventually you've collected answers from every group on the plane (your puzzle input). Each group's answers are separated by a blank line, and within each group, each person's answers are on a single line. For example:

`abc`

`a`
`b`
`c`

`ab`
`ac`

`a`
`a`
`a`
`a`

`b`

This list represents answers from five groups:

- The first group contains one person who answered "yes" to 3 questions: `a`, `b`, and `c`.
- The second group contains three people; combined, they answered "yes" to 3 questions: `a`, `b`, and `c`.
- The third group contains two people; combined, they answered "yes" to 3 questions: `a`, `b`, and `c`.
- The fourth group contains four people; combined, they answered "yes" to only 1 question, `a`.
- The last group contains one person who answered "yes" to only 1 question, `b`.

In this example, the sum of these counts is `3 + 3 + 3 + 1 + 1` = `11`.

For each group, count the number of questions to which anyone answered "yes". 

**What is the sum of those counts?**

```{r load data}
df_raw <- 
  read_lines(file = "20-1206_input_custom_customs.txt") %>% 
  as_tibble()

```

## Exercise 1

```{r wrangle}
df <- 
  df_raw %>% 
  mutate(group = cumsum(. == "") + 1) %>% 
  filter(value != "")

df %>% 
  split(f = as.factor(.$group)) %>% # create a list of each grouping
  map(.f = ~
          # split character strings into individual letters
          mutate(., x = str_split(value, "")) %>% 
          # pull out just our list of vectors of individual letters
          pull(x) %>% 
          # for each list, reduce down to just unique values that appear in any of the lists (union)
          reduce(.f = union) %>% 
          # we really only care about how many, so length()
          length()
        ) %>% 
  # reduce the full list to a single answer by addition
  reduce(.f = `+`)
  
```

The sum for my input for "yes" answers by each group is 6430




## Exercise 2

As you finish the last group's customs declaration, you notice that you misread one word in the instructions:

You don't need to identify the questions to which anyone answered "yes"; you need to identify the questions to which everyone answered "yes"!

Using the same example as above:

`abc`

`a`
`b`
`c`

`ab`
`ac`

`a`
`a`
`a`
`a`

`b`

This list represents answers from five groups:

- In the first group, everyone (all 1 person) answered "yes" to 3 questions: a, b, and c.
- In the second group, there is no question to which everyone answered "yes".
- In the third group, everyone answered yes to only 1 question, a. Since some people did not answer "yes" to b or c, they don't count.
- In the fourth group, everyone answered yes to only 1 question, a.
- In the fifth group, everyone (all 1 person) answered "yes" to 1 question, b.

In this example, the sum of these counts is `3 + 0 + 1 + 1 + 1` = `6`.

For each group, count the number of questions to which everyone answered "yes". What is the sum of those counts?

```{r}
df %>% 
  split(f = as.factor(.$group)) %>% 
  map(.f = ~ 
        # split string into individual characters
        mutate(.,  x = str_split(value, "")) %>% 
        # extract just the split strings, themselves a list
        pull(x) %>% 
        # for each list set of individual characters, reduce to only those they all share (intersect)
        reduce(.f = intersect) %>%
        # count the length (how many unique do they all share)
        length()
  ) %>% 
  # sum up the full list  
  reduce(`+`)
```

The sum of the counts for each group's total of individuals sharing the same "yes" answers is 3125 for my exercise file