---
title: '20-1223_Crab_Cups'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

#disable scientific notation
options(scipen = 999)
```

## Day 23: Crab Cups

The small crab challenges you to a game! The crab is going to mix up some cups, and you have to predict where they'll end up.

The cups will be arranged in a circle and labeled *clockwise* (your puzzle input). For example, if your labeling were `32415`, there would be five cups in the circle; going clockwise around the circle from the first cup, the cups would be labeled `3, 2, 4, 1, 5`, and then back to 3 again.

Before the crab starts, it will designate the first cup in your list as the *current cup*. The crab is then going to do *100 moves*.

Each *move*, the crab does the following actions:

- The crab picks up the *three cups* that are immediately clockwise of the *current cup*. They are removed from the circle; cup spacing is adjusted as necessary to maintain the circle.

- The crab selects a *destination cup*: the cup with a *label* equal to the *current cup's label minus one*. If this would select one of the cups that was just picked up, the crab will keep subtracting one until it finds a cup that wasn't just picked up. If at any point in this process the value goes below the lowest value on any cup's label, it *wraps around* to the highest value on any cup's label instead.

- The crab places the cups it just picked up so that they are *immediately clockwise* of the destination cup. They keep the same order as when they were picked up.

- The crab selects a *new current cup*: the cup which is immediately clockwise of the current cup.

For example, suppose your cup labeling were `389125467`. If the crab were to do merely 10 moves, the following changes would occur:

`-- move 1 --`
`cups: (3) 8  9  1  2  5  4  6  7 `
`pick up: 8, 9, 1`
`destination: 2`

`-- move 2 --`
`cups:  3 (2) 8  9  1  5  4  6  7 `
`pick up: 8, 9, 1`
`destination: 7`

`-- move 3 --`
`cups:  3  2 (5) 4  6  7  8  9  1 `
`pick up: 4, 6, 7`
`destination: 3`

`-- move 4 --`
`cups:  7  2  5 (8) 9  1  3  4  6 `
`pick up: 9, 1, 3`
`destination: 7`

`-- move 5 --`
`cups:  3  2  5  8 (4) 6  7  9  1 `
`pick up: 6, 7, 9`
`destination: 3`

`-- move 6 --`
`cups:  9  2  5  8  4 (1) 3  6  7 `
`pick up: 3, 6, 7`
`destination: 9`

`-- move 7 --`
`cups:  7  2  5  8  4  1 (9) 3  6 `
`pick up: 3, 6, 7`
`destination: 8`

`-- move 8 --`
`cups:  8  3  6  7  4  1  9 (2) 5 `
`pick up: 5, 8, 3`
`destination: 1`

`-- move 9 --`
`cups:  7  4  1  5  8  3  9  2 (6)`
`pick up: 7, 4, 1`
`destination: 5`

`-- move 10 --`
`cups: (5) 7  4  1  8  3  9  2  6 `
`pick up: 7, 4, 1`
`destination: 3`

`-- final --`
`cups:  5 (8) 3  7  4  1  9  2  6 `

In the above example, the cups' values are the labels as they appear moving clockwise around the circle; the *current cup* is marked with `( )`.

After the crab is done, what order will the cups be in? Starting *after the cup labeled 1*, collect the other cups' labels clockwise into a single string with no extra characters; each number except 1 should appear exactly once. In the above example, after 10 moves, the cups clockwise from `1` are labeled` 9, 2, 6, 5`, and so on, producing `92658374`. If the crab were to complete all 100 moves, the order after cup 1 would be `67384529`.

Using your labeling, simulate 100 moves. *What are the labels on the cups after cup 1?*

Your puzzle input is 523764819.

## INPUT

```{r sample input}
sample_raw <- "389125467"

sample <- 
  str_split(sample_raw, "") %>% 
  pluck(1) %>% 
  parse_number()
```


```{r input}
input_raw <- "523764819"

input <- 
  str_split(input_raw, "") %>% 
  pluck(1) %>% 
  parse_number()
```


```{r choose sample or input}
# change below to TRUE to use sample data
use_sample <- FALSE

if (use_sample) input <- sample
```


## EXERCISE 1

```{r exercise 1}
input

crab_cups <- function(all_cups, cup_pos) {
  
  current_cup_value <- all_cups[cup_pos]
  
  remove_pos <- (cup_pos + 1):(cup_pos + 3)
  if (any(remove_pos > 9)) remove_pos[remove_pos > 9] <- remove_pos[remove_pos > 9] - 9
  
  removed <- all_cups[remove_pos]
  
  all_cups <- all_cups[-remove_pos]
  
  
  dest <- which(all_cups < current_cup_value)
  if (length(dest) > 1) dest <- dest[which(all_cups[dest] == max(all_cups[dest]))]
  if (length(dest) == 0) dest <- which(all_cups == max(all_cups))
  
  
  # insert cups back in 
  if (dest == length(all_cups)) {
    res <- c(all_cups, removed)
  } else {
    res <- c(all_cups[1:dest], 
             removed, 
             all_cups[(dest + 1):6])
  }
  
  # shift appropriately, cup that was at the current cup position needs to stay there in the output
  
  cup_loc <- which(res == current_cup_value)
  if (cup_loc == cup_pos) return(res)
  if (cup_loc > cup_pos) {
    diff <- cup_loc - cup_pos
    res <- c(res[(diff + 1):9], res[1:diff])
  } else {
    diff <- cup_pos - cup_loc
    res <- c(tail(res, diff), res)[1:9]
  }
  
  return(res)
  
}

gen_seq_nines <- function(len) {
  n <- (len %/% 9) + 1
  res <- rep(1:9, n)
  res[1:len]
}


get_final_seq_one <- function(nums) {
  one_pos <- which(nums == 1)
  if (one_pos == 1) {
    res <- nums[2:9]
  } else if (one_pos == 0) {
    res <- nums[1:8]
  } else {
    res <- c(nums[(one_pos + 1):9], nums[1:(one_pos - 1)])
  }
  return(str_c(res, collapse = ""))
}


ans1 <- reduce(.x = gen_seq_nines(100), crab_cups, .init = input)

get_final_seq_one(ans1)

```

My answer for exercise 1 is `49576328`



## EXERCISE 2 NARRATIVE

Due to what you can only assume is a mis-translation (you're not exactly fluent in Crab), you are quite surprised when the crab starts arranging many cups in a circle on your raft - *one million* (`1000000`) in total.

Your labeling is still correct for the first few cups; after that, the remaining cups are just numbered in an increasing fashion starting from the number after the highest number in your list and proceeding one by one until one million is reached. (For example, if your labeling were 54321, the cups would be numbered 5, 4, 3, 2, 1, and then start counting up from 6 until one million is reached.) In this way, *every number from one through one million is used exactly once*.

After discovering where you made the mistake in translating Crab Numbers, you realize the small crab isn't going to do merely 100 moves; the crab is going to do *ten million* (*10000000*) moves!

The crab is going to hide your stars - one each - under the two cups that will end up immediately clockwise of cup 1. You can have them if you predict what the labels on those cups will be when the crab is finished.

In the above example (`389125467`), this would be `934001` and then `159792`; multiplying these together produces `149245887792`.

Determine which two cups will end up immediately clockwise of cup 1. 

*What do you get if you multiply their labels together?*

## EXERCISE 2

Code takes a long time to run. Interrupted it after 12 hours before seeing how long it takes to complete. 

```{r exercise 2, eval=FALSE}
million <- 1000000

input <- c(input, (max(input) + 1):million)


gen_seq_millions <- function(len) {
  n <- (len %/% million) + 1
  res <- rep(1:million, n)
  res[1:len]
}


crab_cups_two <- function(all_cups, cup_pos) {
  
  # print(cup_pos)
  
  current_cup_value <- all_cups[cup_pos]
  
  remove_pos <- (cup_pos + 1):(cup_pos + 3)
  if (any(remove_pos > million)) remove_pos[remove_pos > million] <- remove_pos[remove_pos > million] - million
  
  removed <- all_cups[remove_pos]
  
  all_cups <- all_cups[-remove_pos]
  
  
  dest <- which(all_cups < current_cup_value)
  if (length(dest) > 1) dest <- dest[which(all_cups[dest] == max(all_cups[dest]))]
  if (length(dest) == 0) dest <- which(all_cups == max(all_cups))
  
  # insert cups back in 
  if (dest == length(all_cups)) {
    res <- c(all_cups, removed)
  } else {
    res <- c(all_cups[1:dest], 
             removed, 
             all_cups[(dest + 1):(million - 3)])
  }
  
  # shift appropriately, cup that was at the current cup position needs to stay there in the output
  
  cup_loc <- which(res == current_cup_value)
  
  if (cup_loc == cup_pos) return(res)
  
  if (cup_loc > cup_pos) {
    diff <- cup_loc - cup_pos
    res <- c(res[(diff + 1):million], res[1:diff])
  } else {
    diff <- cup_pos - cup_loc
    res <- c(tail(res, diff), res)[1:million]
  }
  
  return(res)
  
}



get_final_seq_two <- function(vec, len = million) {
  pos <- which(vec == 1)
  if (pos == len) return(vec[1:2])
  if (pos == (len - 1)) return(c(vec[len], vec[1]))
  
  return (c(vec[pos + 1], vec[pos + 2]))
}



full_seq <- gen_seq_millions(len = (million * 10))


tictoc::tic("run exercise 2")
for (i in 1:10000) {
  
  if (i == 1) ans2 <- input
  
  if ((i %% 500) == 0) print(i)
  
  ans2 <- crab_cups_two(ans2, i)
  
}
tictoc::toc()


# get_final_seq_two(ans2)
```

## EXERCISE 2 HELP

After trying multiple appraoches, each of which would take days to execute, I looked online for help. 

https://github.com/AdroMine/AdventOfCode2020/blob/main/Day23/solution.R

Essentially, this uses a lookup table to track the next cups in the sequence given a cups values, as opposed to updating an entire vector of length 1 million. Executes part 2 in ~ 15 seconds. 

```{r exercise 2 adromine}

print_list <- function(x, from = "1") {
  output <- rep(NA, length(x))
  output[1] <- as.integer(from)
  for (i in 2:length(x)) {
    output[i] <- x[output[i - 1]]
  }
  output
}


crab_game <- function(input, rounds, part2 = FALSE) {
  
  if (part2) {
    input <- c(input, seq.int(10, 1e6))
  }
  
  n <- length(input)
  
  # create vector to track next cups
  # position in this vector in value of cup you'd like to look up
  # that is, nxt[5] looks up the next cup after the cup with a value of 5
  
  nxt <- integer(n)
  nxt[input[1:(n - 1)]] <- input[-1]
  nxt[input[n]] <- input[1]
  
  cur <- input[[1]]
  
  for (i in seq.int(1, rounds)) {
    
    # choose next three
    pick1 <- nxt[cur]
    pick2 <- nxt[pick1]
    pick3 <- nxt[pick2]
    
    if (cur == 1L) {
      target <- n
    } else {
      target <- (cur - 1L)
    }
    
    # while loop to find where to place cups
    while (target %in% c(pick1, pick2, pick3)) {
      if (target == 1L) {
        target <- n
      } else {
        target <- target - 1L
      } 
    }  
   
    # what's the cup after the ones we removed/shifted
    tmp <- nxt[pick3]
    # the next cup after our current cup will be that value
    nxt[cur] <- tmp
    # inserting the cups after the target will shift the next cup after our third removed cup to the next cup after the target (insertion)
    nxt[pick3] <- nxt[target]
    # next value after the target will be the first from the removed values (insertion)
    nxt[target] <- pick1
    # don't need to update what comes after pick1 or pick2 because those references stay the same as the three are moved in a group!
    
    # next current value update for next loop
    cur <- tmp
     
  }
  
  if (part2) {
    prod(nxt[[1]], 
         nxt[[nxt[[1]]]]
         )
  } else {
    str_c(print_list(nxt)[-1], collapse = "")
  }
  
}

# part 1
tictoc::tic("exercise 1")
crab_game(input, rounds = 100, part2 = FALSE)
tictoc::toc()

# part 2
tictoc::tic("exercise 2")
crab_game(input, rounds = 1e7, part2 = TRUE)
tictoc::toc()


```




