---
title: '20-1222_Crab_Combat'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

#disable scientific notation
options(scipen = 999)
```

## Day 22: Crab Combat

It only takes a few hours of sailing the ocean on a raft for boredom to sink in. Fortunately, you brought a small deck of space cards! You'd like to play a game of *Combat*, and there's even an opponent available: a small crab that climbed aboard your raft before you left.

Fortunately, it doesn't take long to teach the crab the rules.

Before the game starts, split the cards so each player has their own deck (your puzzle input). Then, the game consists of a series of *rounds*: both players draw their top card, and the player with the higher-valued card wins the round. The winner keeps both cards, placing them on the bottom of their own deck so that the winner's card is above the other card. If this causes a player to have all of the cards, they win, and the game ends.

For example, consider the following starting decks:

`Player 1:`
`9`
`2`
`6`
`3`
`1`

`Player 2:`
`5`
`8`
`4`
`7`
`10`

This arrangement means that player 1's deck contains 5 cards, with `9` on top and `1` on the bottom; player 2's deck also contains 5 cards, with `5` on top and `10` on the bottom.

The first round begins with both players drawing the top card of their decks: `9 and 5`. Player 1 has the higher card, so both cards move to the bottom of player 1's deck such that 9 is above 5. In total, it takes 29 rounds before a player has all of the cards:

`-- Round 1 --`
`Player 1's deck: 9, 2, 6, 3, 1`
`Player 2's deck: 5, 8, 4, 7, 10`
`Player 1 plays: 9`
`Player 2 plays: 5`
`Player 1 wins the round!`

`-- Round 2 --`
`Player 1's deck: 2, 6, 3, 1, 9, 5`
`Player 2's deck: 8, 4, 7, 10`
`Player 1 plays: 2`
`Player 2 plays: 8`
`Player 2 wins the round!`

`-- Round 3 --`
`Player 1's deck: 6, 3, 1, 9, 5`
`Player 2's deck: 4, 7, 10, 8, 2`
`Player 1 plays: 6`
`Player 2 plays: 4`
`Player 1 wins the round!`

`-- Round 4 --`
`Player 1's deck: 3, 1, 9, 5, 6, 4`
`Player 2's deck: 7, 10, 8, 2`
`Player 1 plays: 3`
`Player 2 plays: 7`
`Player 2 wins the round!`

`-- Round 5 --`
`Player 1's deck: 1, 9, 5, 6, 4`
`Player 2's deck: 10, 8, 2, 7, 3`
`Player 1 plays: 1`
`Player 2 plays: 10`
`Player 2 wins the round!`

`...several more rounds pass...`

`-- Round 27 --`
`Player 1's deck: 5, 4, 1`
`Player 2's deck: 8, 9, 7, 3, 2, 10, 6`
`Player 1 plays: 5`
`Player 2 plays: 8`
`Player 2 wins the round!`

`-- Round 28 --`
`Player 1's deck: 4, 1`
`Player 2's deck: 9, 7, 3, 2, 10, 6, 8, 5`
`Player 1 plays: 4`
`Player 2 plays: 9`
`Player 2 wins the round!`

`-- Round 29 --`
`Player 1's deck: 1`
`Player 2's deck: 7, 3, 2, 10, 6, 8, 5, 9, 4`
`Player 1 plays: 1`
`Player 2 plays: 7`
`Player 2 wins the round!`


`== Post-game results ==`
`Player 1's deck: `
`Player 2's deck: 3, 2, 10, 6, 8, 5, 9, 4, 7, 1`

Once the game ends, you can calculate the winning player's *score*. The bottom card in their deck is worth the value of the card multiplied by 1, the second-from-the-bottom card is worth the value of the card multiplied by 2, and so on. With 10 cards, the top card is worth the value on the card multiplied by 10. In this example, the winning player's score is:

`   3 * 10`
`+  2 *  9`
`+ 10 *  8`
`+  6 *  7`
`+  8 *  6`
`+  5 *  5`
`+  9 *  4`
`+  4 *  3`
`+  7 *  2`
`+  1 *  1`
`= 306`

So, once the game ends, the winning player's score is `306`.

Play the small crab in a game of Combat using the two decks you just dealt. 

*What is the winning player's score?*

## INPUT

```{r sample input}
sample_raw <- 
  read_file("Player 1:
9
2
6
3
1

Player 2:
5
8
4
7
10")

split_input <- function(.df) {
  str_split(.df, "\\n\\n") %>% 
    pluck(1) %>% 
    str_remove(".* \\d:\\n") %>% 
    str_remove("\\n$") %>% 
    as.list() %>%
    map(~ str_split(., "\\n")[[1]]) %>% 
    map(as.integer)
}

sample <- split_input(sample_raw)


```


```{r input}
input_raw <- read_file("20-1222_input_crab_combat.txt")

input <- split_input(input_raw)

```


```{r choose sample or input}
# change below to TRUE to use sample data
use_sample <- FALSE

if (use_sample) input <- sample
```



## EXERCISE 1

```{r exercise 1}

deal_round <- function(.ls) {
  
  # get first card from each player
  a <- pluck(.ls, 1, 1)
  b <- pluck(.ls, 2, 1)
  
  # remove that card from their hands
  .ls <- map(.ls, tail, -1)
  
  # combination of two cards in descending order that will get added to winner's hand
  comb <- sort(c(a, b), decreasing = TRUE)
  
  # figure out winner
  if (a > b) {
    .ls[[1]] <- c(.ls[[1]], comb)
  } else {
    .ls[[2]] <- c(.ls[[2]], comb)
  }
  
  # return the hands 
  return(.ls)
  
}

solution1 <- input

play_round <- TRUE

while (play_round) {
  
  solution1 <- deal_round(solution1)
  
  have_cards <- 
    map(solution1, ~ length(.) > 0) %>% 
    unlist() %>% 
    sum()
  
  if (have_cards == 1) play_round <- FALSE
  
}

which_deck <- which(unlist(map(solution1, length)) > 0)

solution1 <- pluck(solution1, which_deck)

(solution1 * rev(seq_along(solution1))) %>% sum()

```

Test answer should be `306`

My input answer is `31314`


## EXERCISE 2 NARRATIVE

You lost to the small crab! Fortunately, crabs aren't very good at recursion. To defend your honor as a Raft Captain, you challenge the small crab to a game of *Recursive Combat*.

Recursive Combat still starts by splitting the cards into two decks (you offer to play with the same starting decks as before - it's only fair). Then, the game consists of a series of *rounds* with a few changes:

Before either player deals a card, if there was a previous round in this game that had exactly the same cards in the same order in the same players' decks, the game instantly ends in a win for player 1. Previous rounds from other games are not considered. (This prevents infinite games of Recursive Combat, which everyone agrees is a bad idea.)

Otherwise, this round's cards must be in a new configuration; the players begin the round by each drawing the top card of their deck as normal.

If both players have at least as many cards remaining in their deck as the value of the card they just drew, the winner of the round is determined by playing a new game of Recursive Combat (see below).

Otherwise, at least one player must not have enough cards left in their deck to recurse; the winner of the round is the player with the higher-value card.

As in regular Combat, the winner of the round (even if they won the round by winning a sub-game) takes the two cards dealt at the beginning of the round and places them on the bottom of their own deck (again so that the winner's card is above the other card). Note that the winner's card might be *the lower-valued of the two cards* if they won the round due to winning a sub-game. If collecting cards by winning the round causes a player to have all of the cards, they win, and the game ends.

Here is an example of a small game that would loop forever without the infinite game prevention rule:

`Player 1:`
`43`
`19`

`Player 2:`
`2`
`29`
`14`

During a round of Recursive Combat, if both players have at least as many cards in their own decks as the number on the card they just dealt, the winner of the round is determined by recursing into a sub-game of Recursive Combat. (For example, if player 1 draws the 3 card, and player 2 draws the 7 card, this would occur if player 1 has at least 3 cards left and player 2 has at least 7 cards left, not counting the 3 and 7 cards that were drawn.)

To play a sub-game of Recursive Combat, each player creates a new deck by making a copy of the next cards in their deck (the quantity of cards copied is equal to the number on the card they drew to trigger the sub-game). During this sub-game, the game that triggered it is on hold and completely unaffected; no cards are removed from players' decks to form the sub-game. (For example, if player 1 drew the 3 card, their deck in the sub-game would be copies of the next three cards in their deck.)

After the game, the winning player's score is calculated from the cards they have in their original deck using the same rules as regular Combat. In the above game, the winning player's score is `291`.

Defend your honor as Raft Captain by playing the small crab in a game of Recursive Combat using the same two decks as before. 

*What is the winning player's score?*

## EXERCISE 2

Eval set to false here because this approach didn't work.

Started again in code chunk `attempt 2` below. 

```{r exercise 2, eval=FALSE}
solution2 <- input

prior_hands <- list(NA)

input_test <- function(.l) sum(unlist(map(.l, ~ length(.) > 0))) > 1

n <- 0

while (input_test(solution2)) {
  
  n <- n + 1
  
  # if we've seen this hand before, player A wins by default to avoid infinite loops
  
  if (reduce(map(prior_hands, ~ identical(.x, solution2)), `+`) > 0) {
    
    # deal top two and add them to player1 with player's one card before player2's
    solution2[[1]] <- 
      c(solution2[[1]], 
        c(pluck(solution2, 1, 1), 
          pluck(solution2, 2, 1)))
    
    solution2 <- map(solution2, tail, -1)
    
    next
  }
  
  # if we haven't seen this hand before, add it
  prior_hands[[n]] <- solution2  
  
  
  # FUNCTION
  
  deal_winner <- function(.ls) {
    
    # if only one hand has cards, it is the winner
    has_cards <- 
      unlist(map(.ls, ~ length(.) > 0))
    
    if (sum(has_cards) == 1) {
      
      end <- which(has_cards == TRUE)
      
      if (end == 1) {
        res <- "a"
      } else {
        res <- "b"
      }
      
      return(res)
      
    }
    
    # deal cards
    # if one (or both) player's deal a card whose value is higher than the amount of cards they have left in their deck, we play a normal round where the highest card wins
    
    cards <- c(pluck(.ls, 1, 1), 
               pluck(.ls, 2, 1))
    
    if (cards[1] > (length(.ls[[1]]) - 1) ||
        cards[2] > (length(.ls[[2]]) - 1) 
    ) {
      
      if (cards[1] > cards[2]) {
        
        # .ls[[1]] <- c(.ls[[1]], 
        #               c(pluck(.ls, 1, 1), 
        #                 pluck(.ls, 2, 1)))
        
        return("a")
        
      } else {
        
        # .ls[[2]] <- c(.ls[[2]], 
        #               c(pluck(.ls, 2, 1), 
        #                 pluck(.ls, 1, 1)))
        
        return("b")
        
      }
      
      
      # .ls <- map(.ls, tail, -1)
      
      # res <- deal_winner(.ls)
      
      
    } else {
      
      # if above condition isn't met, then both players have enough cards to enter into recursive combat
      
      rec_sub_a <- 
        tail(.ls[[1]], -1) %>% # remove top card (already dealt)
        .[1:cards[1]] # take top n matching card value
      rec_sub_b <- 
        tail(.ls[[2]], -1) %>% 
        .[1:cards[2]]
      
      rec_sub <- list(rec_sub_a, rec_sub_b)
      
      res <- deal_winner(rec_sub)
      
    }
    
    return(res)
    
  } # end function
  
  
  round_winner <- deal_winner(solution2)
  
  if (round_winner == "a") {
    solution2[[1]] <- 
      c(solution2[[1]], 
        c(pluck(solution2, 1, 1),
          pluck(solution2, 2, 1)))
    
    solution2 <- map(solution2, tail, -1)
    
  } else {
    solution2[[2]] <- 
      c(solution2[[2]],
        c(pluck(solution2, 2, 1),
          pluck(solution2, 1, 1)))
    
    solution2 <- map(solution2, tail, -1)
    
  }
  
  
} # end while loop

solution2

```


## EXERCISE 2 ATTEMPT 2

```{r exercise 2 attempt 2, eval=FALSE}
# options(expressions = 15000)


solution2 <- input

# prior_hands <- list(NA)

play_combat <- function(.ls, prior_hands = list(list("a", "b")), auto_win = FALSE, ...) {
  
  # if we've seen hand before, player 1 auto_wins
  if (
    map(prior_hands, ~ identical(.x, .ls)) %>% 
    reduce(., `+`) %>% 
    `>` (., 0)
  ) auto_win <- TRUE
  
  
  if (auto_win == FALSE) { 
    prior_hands <- c(prior_hands, list(.ls))
  }  
  
  # print(prior_hands)
  
  cards <- c(pluck(.ls, 1, 1), pluck(.ls, 2, 1))
  
  cards_len <- length(cards) # if 1, one of them's run out of cards
  
  if (cards_len == 1) return(.ls)
  
  
  # remove cards dealt
  .ls <- map(.ls, tail, -1)
  
  # if we have the cards available, enter recursive combat
  if (auto_win == FALSE &&
      cards[1] <= (length(.ls[[1]])) &&    
      cards[2] <= (length(.ls[[2]]))
  ) {
    
    recurse <- .ls
    
    # take top n according to card number
    recurse[[1]] <- head(.ls[[1]], cards[1])
    recurse[[2]] <- head(.ls[[2]], cards[2])
    
    
    .ls <- play_combat(recurse)
    
  }
  
  
  
  # check winner and add cards accordingly
  if (auto_win || 
      cards[1] > cards[2]) {
    .ls[[1]] <- c(.ls[[1]], cards)
  } else {
    .ls[[2]] <- c(.ls[[2]], rev(cards))
  }
  
  
  loop_check <- function(.ls) {
    map(.ls, ~ length(.) > 0) %>%
      reduce(., sum) %>%
      `>` (., 1)
  }
  
  while (loop_check(.ls)) {
    
    .ls <- play_combat(.ls, prior_hands = prior_hands)
    
  }
  
  return(.ls)
}



# check_play <- TRUE
# 
# while (check_play) {
#   
#   solution2 <- play_combat(solution2)
#   
#   check_play <-
#     map(solution2, ~ length(.) > 0) %>%
#     reduce(., sum) %>%
#     `>` (., 1)
#   
# }

play_combat(input)


# accumulate(1:5, play_combat, .init = split_input(input_raw))


# play_combat(.ls = list(c(4,9,8,5,2), c(3,10,1,7,6)))

# debugonce(play_combat)


# solution2_check <- 
#   which(unlist(map(solution2, ~ length(.) > 0)))
# 
# solution2_hand <- solution2[[solution2_check]]
# 
# (solution2_hand * 
#     rev(seq_along(solution2_hand)) ) %>% 
#   sum()


```


## EXERCISE 2 ATTEMPT 3

Use function just to return resolve to "a" or "b"?

Also fails

```{r ex2 attempt 3, eval=FALSE}
solution2 <- input


combat <- function(.ls, 
                   prior_hands = list(list("a", "b")), 
                   auto_win = FALSE) {
  
  names(.ls) <- NULL
  
  # if (.ls %in% c("a", "b")) return(.ls)
  
  # if one of them doesn't have cards, return the winner
  card_check_len <- function(.x) unlist(map(.x, ~ length(.) > 0))
  
  if (sum(card_check_len(.ls)) == 1) {
    
    winner <- which(card_check_len(.ls) == TRUE)
    
    if (winner == 1) {
      names(.ls) <- "a"
      return(.ls)
    } else {
      names(.ls) <- "b"
      return(.ls)
    }
    
  }
  
  # deal cards
  cards <- c(pluck(.ls, 1, 1), 
             pluck(.ls, 2, 1))
  
  # remove dealt cards
  .ls <- map(.ls, tail, -1)
  
  # check for auto-win
  auto_win <- reduce(map(prior_hands, ~ identical(.x, .ls)), sum) > 0
  
  
  # add this hand to prior_hands
  prior_hands <- c(prior_hands, list(.ls))
  
  
  # check for recursion needed
  recurse_check <- all((unlist(map(.ls, length)) - cards) >= 0 )
  
  if (recurse_check == TRUE && auto_win == FALSE) {
    # take n number of cards from each
    .ls <- map2(.ls, cards, ~ head(x = .x, n = .y))
    
    .ls <- combat(.ls)
    
  }
  
  # if auto_win, player a wins by default
  if (auto_win || (!is.null(names(.ls)) && names(.ls)[[1]] == "a")) {
    .ls[[1]] <- c(.ls[[1]], cards)
    
  } else if (!is.null(names(.ls)) && names(.ls)[[1]] == "b") {
    .ls[[2]] <- c(.ls[[2]], rev(cards))
    
  } else if (cards[1] > cards[2]) {
    .ls[[1]] <- c(.ls[[1]], cards)
    
  } else {
    .ls[[2]] <- c(.ls[[2]], rev(cards))
    
  }
  
  while (sum(card_check_len(.ls)) > 1) {
    
    .ls <- combat(.ls, prior_hands = prior_hands)
    
    print(.ls)
    
  }
  
  winner <- which(card_check_len(.ls) == TRUE)
  
  if (winner == 1) {
    names(.ls) <- "a"
  } else {
    names(.ls) <- "b"
  }
  
  return(.ls)
  
}

# debugonce(combat)

combat(solution2)

```



## EXERCISE 2 ATTEMPT 4

Now looking for help online -- found posted to AoC subreddit for RLang by user u/SpirePicking. 

Their solution is below.  I then worked through it step by step with my own (tidyverse) edits to see if I could understand it and get it to work. 

```{r, eval=FALSE}
library(tidyr)

input <- readLines("20-1222_input_crab_combat.txt")

decks <- list("p1" = input[2:(which(input=="")-1)] %>% as.numeric %>% as.list,
              "p2" = input[(which(input=="")+2):length(input)] %>% as.numeric %>% as.list)

update_decks <- function(decks, winning_player) {
  losing_player <- winning_player %% 2 + 1
  winning_deck <- decks[[winning_player]]
  losing_deck <- decks[[losing_player]]
  
  upd_winning_deck <- append(winning_deck, c(winning_deck[1], losing_deck[1]))[-1]
  upd_losing_deck <- losing_deck[-1]
  assign(paste0("upd_deck_p", winning_player), upd_winning_deck)
  assign(paste0("upd_deck_p", losing_player), upd_losing_deck)
  upd_decks <- list("p1" = upd_deck_p1, "p2" = upd_deck_p2)
  
  return(upd_decks)
}

play_combat <- function(decks) {
  while (length(decks[["p1"]]) > 0 & length(decks[["p2"]] > 0)) {
    winning_player <- ifelse(decks[["p1"]][[1]] > decks[["p2"]][[1]], 1, 2)
    decks <- update_decks(decks, winning_player)
  }
  return(list(winning_player, decks[[winning_player]]))
}

play_recursive_combat <- function(decks) {
  
  is_already_seen <- function(deck, configurations_seen){
    if (length(configurations_seen) == 0) {return(0)}
    return(max(sapply(configurations_seen, function(x){identical(deck, x)}))) 
  }
  
  configurations_seen <- list()
  
  while (length(decks[["p1"]]) > 0 & length(decks[["p2"]]) > 0) {
    if(is_already_seen(decks[["p1"]], configurations_seen)) {
      return(list(c(1), decks[["p1"]]))
    }
    configurations_seen <- append(configurations_seen, list(decks[["p1"]]))
    
    is_next_round_recusive <- length(decks[["p1"]][-1]) >= decks[["p1"]][1] & length(decks[["p2"]][-1]) >= decks[["p2"]][1]
    if(is_next_round_recusive){
      winning_player <- play_recursive_combat(list("p1" = decks[["p1"]][2:(1+decks[["p1"]][[1]])],
                                                   "p2" = decks[["p2"]][2:(1+decks[["p2"]][[1]])])
      )[[1]]
    } else { 
      winning_player <- ifelse(decks[["p1"]][[1]] > decks[["p2"]][[1]], 1, 2)
    }
    decks <- update_decks(decks, winning_player)
  }
  return(list(winning_player, decks[[winning_player]]))
}

ans1 <- function(decks) {
  combat_results <- play_combat(decks)
  return(sum(length(combat_results[[2]]):1 * unlist(combat_results[[2]])))
}

ans2 <- function(decks) {
  recursive_combat_results <- play_recursive_combat(decks)
  return(sum(length(recursive_combat_results[[2]]):1 * unlist(recursive_combat_results[[2]])))
}

ans1(decks)
ans2(decks)
```


My code works for the sample but not the real data.  I give up for now. 

```{r}

update_decks <- function(decks, winner) {
  cards <- c(pluck(decks, 1, 1), 
             pluck(decks, 2, 1))
  decks <- map(decks, tail, -1)
  if (winner == 1) {
    decks[[1]] <- c(decks[[1]], cards[1], cards[2])
  } else {
    decks[[2]] <- c(decks[[2]], cards[2], cards[1])
  }
  
  return(decks)
}

rec_combat <- function(decks, auto_win = FALSE) {
  
  config_seen <- function(decks, configs) {
    if (length(configs) == 0) return(0)
    return(map(configs, ~ identical(decks, .x)) %>% unlist() %>% as.numeric() %>% max())
  }
  
  configs <- list()
  
  check_deck_len <- function(decks) {
    map(decks, ~ length(.) > 0) %>% 
      unlist() %>% 
      min()
  }
  
  while(check_deck_len(decks) > 0) {
    
    if (config_seen(decks, configs)) {
      auto_win <- TRUE
    }
    
    configs <- append(configs, list(decks))
    
    need_recursive <- (length(decks[[1]]) - 1) >= pluck(decks, 1, 1) && (length(decks[[2]] - 1) >= pluck(decks, 2, 1))
    
    if (need_recursive && auto_win == FALSE) {
      rec_cards <- c(pluck(decks, 1, 1), pluck(decks, 2, 1))
      rec_deck <- map(decks, tail, -1) 
      rec_deck <- map2(rec_deck, rec_cards, ~ head(.x, .y))
        
      winning_player <- (rec_combat(decks = rec_deck))[[1]]
    } else { 
      winning_player <- case_when(auto_win == TRUE ~ 1,
                                  pluck(decks, 1, 1) > pluck(decks, 2, 1) ~ 1,
                                  TRUE ~ 2)
    }
    
    decks <- update_decks(decks, winning_player)
    
  }
  
  winning_player <- which(unlist(map(decks, ~ length(.) > 0)) == TRUE)
  
  
  return(list(winning_player, decks[[winning_player]]))
}



solution2 <- rec_combat(input) 

sum(solution2[[2]] * (rev(seq_along(solution2[[2]]))))

# debugonce(rec_combat)
```

