---
title: '20-1223_Crab_Cups'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

#disable scientific notation
options(scipen = 999)
```

## Day 23: Crab Cups

The small crab challenges you to a game! The crab is going to mix up some cups, and you have to predict where they'll end up.

The cups will be arranged in a circle and labeled *clockwise* (your puzzle input). For example, if your labeling were `32415`, there would be five cups in the circle; going clockwise around the circle from the first cup, the cups would be labeled `3, 2, 4, 1, 5`, and then back to 3 again.

Before the crab starts, it will designate the first cup in your list as the *current cup*. The crab is then going to do *100 moves*.

Each *move*, the crab does the following actions:

- The crab picks up the *three cups* that are immediately clockwise of the *current cup*. They are removed from the circle; cup spacing is adjusted as necessary to maintain the circle.

- The crab selects a *destination cup*: the cup with a *label* equal to the *current cup's label minus one*. If this would select one of the cups that was just picked up, the crab will keep subtracting one until it finds a cup that wasn't just picked up. If at any point in this process the value goes below the lowest value on any cup's label, it *wraps around* to the highest value on any cup's label instead.

- The crab places the cups it just picked up so that they are *immediately clockwise* of the destination cup. They keep the same order as when they were picked up.

- The crab selects a *new current cup*: the cup which is immediately clockwise of the current cup.

For example, suppose your cup labeling were `389125467`. If the crab were to do merely 10 moves, the following changes would occur:

`-- move 1 --`
`cups: (3) 8  9  1  2  5  4  6  7 `
`pick up: 8, 9, 1`
`destination: 2`

`-- move 2 --`
`cups:  3 (2) 8  9  1  5  4  6  7 `
`pick up: 8, 9, 1`
`destination: 7`

`-- move 3 --`
`cups:  3  2 (5) 4  6  7  8  9  1 `
`pick up: 4, 6, 7`
`destination: 3`

`-- move 4 --`
`cups:  7  2  5 (8) 9  1  3  4  6 `
`pick up: 9, 1, 3`
`destination: 7`

`-- move 5 --`
`cups:  3  2  5  8 (4) 6  7  9  1 `
`pick up: 6, 7, 9`
`destination: 3`

`-- move 6 --`
`cups:  9  2  5  8  4 (1) 3  6  7 `
`pick up: 3, 6, 7`
`destination: 9`

`-- move 7 --`
`cups:  7  2  5  8  4  1 (9) 3  6 `
`pick up: 3, 6, 7`
`destination: 8`

`-- move 8 --`
`cups:  8  3  6  7  4  1  9 (2) 5 `
`pick up: 5, 8, 3`
`destination: 1`

`-- move 9 --`
`cups:  7  4  1  5  8  3  9  2 (6)`
`pick up: 7, 4, 1`
`destination: 5`

`-- move 10 --`
`cups: (5) 7  4  1  8  3  9  2  6 `
`pick up: 7, 4, 1`
`destination: 3`

`-- final --`
`cups:  5 (8) 3  7  4  1  9  2  6 `

In the above example, the cups' values are the labels as they appear moving clockwise around the circle; the *current cup* is marked with `( )`.

After the crab is done, what order will the cups be in? Starting *after the cup labeled 1*, collect the other cups' labels clockwise into a single string with no extra characters; each number except 1 should appear exactly once. In the above example, after 10 moves, the cups clockwise from `1` are labeled` 9, 2, 6, 5`, and so on, producing `92658374`. If the crab were to complete all 100 moves, the order after cup 1 would be `67384529`.

Using your labeling, simulate 100 moves. *What are the labels on the cups after cup 1?*

Your puzzle input is 523764819.

## INPUT

```{r sample input}
sample_raw <- "389125467"

sample <- 
  str_split(sample_raw, "") %>% 
  pluck(1) %>% 
  parse_number()
```


```{r input}
input_raw <- "523764819"

input <- 
  str_split(input_raw, "") %>% 
  pluck(1) %>% 
  parse_number()
```


```{r choose sample or input}
# change below to TRUE to use sample data
use_sample <- FALSE

if (use_sample) input <- sample
```


## EXERCISE 1

```{r exercise 1}
input

crab_cups <- function(all_cups, cup_pos) {
  
  current_cup_value <- all_cups[cup_pos]
  
  remove_pos <- (cup_pos + 1):(cup_pos + 3)
  if (any(remove_pos > 9)) remove_pos[remove_pos > 9] <- remove_pos[remove_pos > 9] - 9
  
  removed <- all_cups[remove_pos]
  
  all_cups <- all_cups[-remove_pos]
  
  
  dest <- which(all_cups < current_cup_value)
  if (length(dest) > 1) dest <- dest[which(all_cups[dest] == max(all_cups[dest]))]
  if (length(dest) == 0) dest <- which(all_cups == max(all_cups))
  
  
  # insert cups back in 
  if (dest == length(all_cups)) {
    res <- c(all_cups, removed)
  } else {
    res <- c(all_cups[1:dest], 
             removed, 
             all_cups[(dest + 1):6])
  }
  
  # shift appropriately, cup that was at the current cup position needs to stay there in the output
  
  cup_loc <- which(res == current_cup_value)
  if (cup_loc == cup_pos) return(res)
  if (cup_loc > cup_pos) {
    diff <- cup_loc - cup_pos
    res <- c(res[(diff + 1):9], res[1:diff])
  } else {
    diff <- cup_pos - cup_loc
    res <- c(tail(res, diff), res)[1:9]
  }
  
  return(res)
  
}

gen_seq_nines <- function(len) {
  n <- (len %/% 9) + 1
  res <- rep(1:9, n)
  res[1:len]
}


get_final_seq <- function(nums) {
  one_pos <- which(nums == 1)
  if (one_pos == 1) {
    res <- nums[2:9]
  } else if (one_pos == 0) {
    res <- nums[1:8]
  } else {
    res <- c(nums[(one_pos + 1):9], nums[1:(one_pos - 1)])
  }
  return(str_c(res, collapse = ""))
}


ans1 <- reduce(.x = gen_seq_nines(100), crab_cups, .init = input)

get_final_seq(ans1)

```

My answer for exercise 1 is `49576328`


## EXERCISE 2 NARRATIVE


## EXERCISE 2

```{r}

```

