---
title: '20-1211_Seating_System'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

#disable scientific notation
options(scipen = 999)
```

## Day 11: Seating System 

Your plane lands with plenty of time to spare. The final leg of your journey is a ferry that goes directly to the tropical island where you can finally start your vacation. As you reach the waiting area to board the ferry, you realize you're so early, nobody else has even arrived yet!

By modeling the process people use to choose (or abandon) their seat in the waiting area, you're pretty sure you can predict the best place to sit. You make a quick map of the seat layout (your puzzle input).

The seat layout fits neatly on a grid. Each position is either floor (.), an empty seat (L), or an occupied seat (#). For example, the initial seat layout might look like this:

`L.LL.LL.LL`
`LLLLLLL.LL`
`L.L.L..L..`
`LLLL.LL.LL`
`L.LL.LL.LL`
`L.LLLLL.LL`
`..L.L.....`
`LLLLLLLLLL`
`L.LLLLLL.L`
`L.LLLLL.LL`

Now, you just need to model the people who will be arriving shortly. Fortunately, people are entirely predictable and always follow a simple set of rules. All decisions are based on the *number of occupied seats* adjacent to a given seat (one of the eight positions immediately up, down, left, right, or diagonal from the seat). The following rules are applied to every seat simultaneously:

- If a seat is *empty* (L) and there are *no* occupied seats adjacent to it, the seat becomes *occupied.*
- If a seat is *occupied* (#) and *four or more seats adjacent* to it are also occupied, the seat becomes *empty*.
- Otherwise, the seat's state does not change.

Floor (.) never changes; seats don't move, and nobody sits on the floor.

After one round of these rules, every seat in the example layout becomes occupied:

`#.##.##.##`
`#######.##`
`#.#.#..#..`
`####.##.##`
`#.##.##.##`
`#.#####.##`
`..#.#.....`
`##########`
`#.######.#`
`#.#####.##`

After a second round, the seats with four or more occupied adjacent seats become empty again:

`#.LL.L#.##`
`#LLLLLL.L#`
`L.L.L..L..`
`#LLL.LL.L#`
`#.LL.LL.LL`
`#.LLLL#.##`
`..L.L.....`
`#LLLLLLLL#`
`#.LLLLLL.L`
`#.#LLLL.##`

This process continues for three more rounds:

`#.##.L#.##`
`#L###LL.L#`
`L.#.#..#..`
`#L##.##.L#`
`#.##.LL.LL`
`#.###L#.##`
`..#.#.....`
`#L######L#`
`#.LL###L.L`
`#.#L###.##`

`#.#L.L#.##`
`#LLL#LL.L#`
`L.L.L..#..`
`#LLL.##.L#`
`#.LL.LL.LL`
`#.LL#L#.##`
`..L.L.....`
`#L#LLLL#L#`
`#.LLLLLL.L`
`#.#L#L#.##`

`#.#L.L#.##`
`#LLL#LL.L#`
`L.#.L..#..`
`#L##.##.L#`
`#.#L.LL.LL`
`#.#L#L#.##`
`..L.L.....`
`#L#L##L#L#`
`#.LLLLLL.L`
`#.#L#L#.##`

At this point, something interesting happens: the chaos stabilizes and further applications of these rules cause no seats to change state! Once people stop moving around, you count *37* occupied seats.

Simulate your seating area by applying the seating rules repeatedly until no seats change state. 

*How many seats end up occupied?*

```{r}
input_raw <- read_lines("20-1211_input_seating_system.txt")

input <- 
  input_raw %>% 
  map(., ~ as.vector(x = str_split(., "", simplify = TRUE)))
```


## EXERCISE 1

```{r exercise 1}
n <- 0

seat_assign <- function(df) {
  
  args <-   
    list(a = df, 
         b = lag(df), 
         c = lead(df))
  
  # results
  intermediate <- vector("list", length(df))
  
  for (i in 1:length(df)) {
    intermediate[[i]] <- rep(".", length(df[[1]]))
  }
  
  
  for (i in seq_along(args$a)) {
    
    maxspot <- length(args$a[[i]])
    
    for (j in seq_along(args$a[[i]]) ) {
      
      spot <- args$a[[i]][j]
      
      if (spot == ".") next # floor spots never change
      
      occupied <- 
        sum(
          c(
            replace_na(args$b[[i]][(j-1):(j+1)], "^"), 
            replace_na(args$a[[i]][(j-1)], "^"), 
            replace_na(args$a[[i]][(j+1)], "^"), 
            replace_na(args$c[[i]][(j-1):(j+1)], "^") 
          ) == "#"
        )
      
      # if a seat is empty and there are no occupied seats adjacent, it becomes occupied
      if (spot == "L" && occupied == 0) {
        
        intermediate[[i]][j] <- "#"
        
        # if a seat is occupied and four or more seats adjacent become occupied it becomes empty
      } else if (spot == "#" && occupied >= 4) {
        
        intermediate[[i]][j] <- "L"
        
        # otherwise, the seat stays the same
      } else {
        
        intermediate[[i]][j] <- spot
        
      }
      
    } # end inner loop (j)
    
  } # end outer loop (i)
  
  n <<- n + 1
  print(n)
  
  if (identical(intermediate, df)) {
    return(intermediate)
  } else {
    seat_assign(df = intermediate)
  }
  
  
} #end function


res <- seat_assign(input)


exercise1 <- reduce(map(res, ~ sum(. == "#")), `+`)

exercise1
```

`2338` seats end up occupied

## EXERCISE 2

I did exercise 1 using lists and it wasn't super easy

I'll try exercise 2 using a character matrix

```{r}
# VERY IMPORTANT TO DO THE "byrow = TRUE", else it will create the matrix filling cols first
mat <- reduce(input, c) %>% matrix(., nrow = length(input), byrow = TRUE)

reduce_eval <- function(string) {
  str_c(string, collapse = "") %>% 
    str_match(., "(L|#)") %>% 
    .[[1]] %>% 
    replace_na(., "L")  == "#"
}

n2 <- 0

seat_assign_diagonal <- function(df) {
  
  # top left across, then down
  # [row, col]
  nr <- nrow(df)
  nc <- ncol(df)
  
  intermediate <- 
    rep("^", nr * nc) %>% 
    matrix(ncol = nc)
  
  seat_symb <- c("L", "#", ".")
  
  for (i in 1:nr) {
    
    for (j in 1:nc) {
      
      seat <- df[i,j]
      
      res <- vector("integer", 8)
      
      # up (1)
      if (i == 1) { # first row
        res[1] <- 0
      } else {
        res[1] <- if_else(df[(i-1):1, j] %>% reduce_eval(), 1, 0)
      }
      
      # down (5)
      if (i == nr) { # last row
        res[5] <- 0
      } else {
        res[5] <- if_else(df[(i+1):nr, j] %>% reduce_eval(), 1, 0)
      }
      
      # right (3)
      if (j == nc) {
        res[3] <- 0
      } else {
        res[3] <- if_else(df[i, (j+1):nc] %>% reduce_eval(), 1, 0)
      }
      
      # left (7)
      if (j == 1) {
        res[7] <- 0
      } else {
        res[7] <- if_else(df[i, (j-1):1] %>% reduce_eval(), 1, 0)
      }
      
      # up/right diagonal (2)
      if (i == 1 || j == nc) {
        res[2] <- 0
      } else {
        res[2] <- 
          if_else(df[1:(i-1), (j+1):nc, drop = FALSE] %>% 
                    .[nrow(.):1, , drop = FALSE] %>% #rotate
                    diag() %>% 
                    reduce_eval(), 1, 0)
      } 
      
      # down/right diagonal (4)
      if (i == nr || j == nc) {
        res[4] <- 0
      } else {
        res[4] <- 
          if_else(df[(i+1):nr, (j+1):nc, drop = FALSE] %>% 
                    diag() %>%  # no rotation need this direction
                    reduce_eval(), 1, 0)
      }
      
      
      # down/left diagonal (6)
      if (i == nr || j == 1) {
        res[6] <- 0
      } else {
        res[6] <- 
          if_else(df[(i+1):nr, 1:(j-1), drop = FALSE] %>% 
                    .[, ncol(.):1, drop = FALSE] %>% #rotate
                    diag() %>% 
                    reduce_eval(), 1, 0)
      }
      
      
      # up/left diagonal (8)
      if (i == 1 || j == 1) {
        res[8] <- 0
      } else {
        res[8] <- 
          if_else(df[1:(i-1), 1:(j-1), drop = FALSE] %>% 
                    .[nrow(.):1, ncol(.):1, drop = FALSE] %>% #rotate
                    diag() %>% 
                    reduce_eval(), 1, 0)
      }
      
      # seat evaluation
      if (seat == ".") {
        intermediate[i,j] <- "."
        
      } else if (seat == "L" && sum(res) == 0) {
        intermediate[i,j] <- "#"
        
      } else if (seat == "#" && sum(res) >= 5) {
        intermediate[i,j] <- "L"
        
      } else {
        intermediate[i,j] <- seat
      }
      
      
    } # end column loop (j)
    
  } # end row loop (i)
  
  n2 <<- n2 + 1
  print(n2)
  
  if (identical(intermediate, df)) {
    return(intermediate)
  } else {
    seat_assign_diagonal(intermediate)
  }
  
} # end function

```

```{r exercise 2 execute}
res2 <- seat_assign_diagonal(mat)

sum(res2 == "#")


```

I got `2088`, but that's *incorrect*


Turns out, I created the intial matrix incorrectly (hadn't had `byrows = TRUE` set).
When I create this matrix correctly it returns `2134` which is, in fact, the right answer.
That said, this code takes a LONG time to run. More efficient solution below. 



## Alternative solution EXERCISE 1

```{r alt exercise 1}
mat <- reduce(input, c) %>% matrix(., nrow = length(input), byrow = TRUE)


find_neighbor_immediate <- function(m, x, y) {
  
  tmp_mat <- m
  
  tmp_mat[x, y] <- ""
  
  tmp_mat[seq(max(x-1, 1), min(x+1, nrow(m))), 
          seq(max(y-1, 1), min(y+1, ncol(m)))] 
  
}

# m_start <- mat

n1 <- 0

exercise1 <- function(m_start) {
  
  n1 <<- n1 + 1
  
  print(n1)
  
  m_replace <- m_start
  
  for (i in seq_len(nrow(m_start))) {
    
    for (j in seq_len(ncol(m_start))) {
      
      seat <- m_start[i, j]
      
      if (seat == ".") next
      
      nbors <- find_neighbor_immediate(m_start, i, j) 
      nbors <- sum(nbors == "#")
      
      # if seat is empty and no adjacent seats occupied, seat becomes occupied
      if (seat == "L" && nbors == 0) m_replace[i, j] <- "#"
      
      # if seat is occupied and four or more adjacent seats are also occupied, seat becomes empty
      if (seat == "#" && nbors >= 4) m_replace[i, j] <- "L"
      
    } # col loop
    
  } # row loop
  
  if (identical(m_start, m_replace)) {
    return(m_replace)
  } else {
    exercise1(m_start = m_replace)
  }
  
  
} # end function


#res1

res1 <- exercise1(m_start = mat)

sum(res1 == "#")
```

`2338` seats occupied 

## Alternate Solution EXERCISE 2

```{r alt exercise 2}
mat <- reduce(input, c) %>% matrix(., nrow = length(input), byrow = TRUE)


find_neighbor_extended <- function(m, x, y, xadd, yadd) {
  
  if (xadd == 0 && yadd == 0) stop(simpleError("Either xadd or yadd must not be zero"))
  
  res <- "."
  
  # step in matrix given inputs
  i <- x + xadd
  j <- y + yadd
  
  while (i >= 1 && j >= 1 && i <= nrow(m) && j <= ncol(m)) {
    
    value <- m[i,j]
    
    if (value != ".") {
      res <- value
      break
    } 
    
    i <- i + xadd
    j <- j + yadd
    
  }
  res
}


find_neighbor_extended_all <- function(m, x, y) {
  
  north     <- find_neighbor_extended(m, x, y, xadd = -1, yadd = 0)
  northeast <- find_neighbor_extended(m, x, y, xadd = -1, yadd = 1)
  east      <- find_neighbor_extended(m, x, y, xadd = 0 , yadd = 1)
  southeast <- find_neighbor_extended(m, x, y, xadd = 1 , yadd = 1)
  south     <- find_neighbor_extended(m, x, y, xadd = 1 , yadd = 0)
  southwest <- find_neighbor_extended(m, x, y, xadd = 1 , yadd = -1)
  west      <- find_neighbor_extended(m, x, y, xadd = 0 , yadd = -1)
  northwest <- find_neighbor_extended(m, x, y, xadd = -1, yadd = -1)
  
  c(north, northeast, east, southeast, south, southwest, west, northwest)
}



n2 <- 0


exercise2 <- function(m_start) {
  
  n2 <<- n2 + 1
  print(n2)
  
  m_replace <- m_start
  
  for (i in seq_len(nrow(m_start))) {
    
    for (j in seq_len(ncol(m_start))) {
      
      pos <- m_start[i, j]
      
      if (pos == ".") next
      
      nbors <- find_neighbor_extended_all(m_start, i, j)
      nbors <- sum(nbors == "#")
      
      # if seat is empty and it can't "see" any other taken seats, seat will become occupied
      if (pos == "L" && nbors == 0) m_replace[i, j] <- "#"
      
      # if seat is taken and they can "see" at least 5 other taken seats, seat will vacate
      if (pos == "#" && nbors >= 5) m_replace[i, j] <- "L"
      
    } # col loop
    
  } # row loop
  
  if (identical(m_start, m_replace)) {
    return(m_replace)
  } else {
    exercise2(m_replace)
  }
  
} # end function

res2 <- exercise2(mat)

res2 <- sum(res2 == "#")

res2

```

`2134` seats taken