---
title: '20-1210_Adapter_Array'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)

  #disable scientific notation
options(scipen = 999)
```

## Day 10: Adapter Array

Patched into the aircraft's data port, you discover weather forecasts of a massive tropical storm. Before you can figure out whether it will impact your vacation plans, however, your device suddenly turns off!

Its battery is dead.

You'll need to plug it in. There's only one problem: the charging outlet near your seat produces the wrong number of jolts. Always prepared, you make a list of all of the joltage adapters in your bag.

Each of your joltage adapters is rated for a specific *output joltage* (your puzzle input). Any given adapter can take an input `1`, `2`, or `3` jolts *lower* than its rating and still produce its rated output joltage.

In addition, your device has a built-in joltage adapter rated for *3 jolts higher* than the highest-rated adapter in your bag. (If your adapter list were `3`, `9`, and `6`, your device's built-in adapter would be rated for `12` jolts.)

Treat the charging outlet near your seat as having an effective joltage rating of 0.

Since you have some time to kill, you might as well test all of your adapters. Wouldn't want to get to your resort and realize you can't even charge your device!

If you *use every adapter in your bag* at once, what is the distribution of joltage differences between the charging outlet, the adapters, and your device?

For example, suppose that in your bag, you have adapters with the following joltage ratings:

`16`
`10`
`15`
`5`
`1`
`11`
`7`
`19`
`6`
`12`
`4`

With these adapters, your device's built-in joltage adapter would be rated for 19 + 3 = 22 jolts, 3 higher than the highest-rated adapter.

Because adapters can only connect to a source 1-3 jolts lower than its rating, in order to use every adapter, you'd need to choose them like this:

- The charging outlet has an effective rating of `0` jolts, so the only adapters that could connect to it directly would need to have a joltage rating of `1`, `2,` or `3` jolts. Of these, only one you have is an adapter rated 1 jolt (difference of 1).
- From your 1-jolt rated adapter, the only choice is your 4-jolt rated adapter (difference of 3).
- From the 4-jolt rated adapter, the adapters rated 5, 6, or 7 are valid choices. However, in order to not skip any adapters, you have to pick the adapter rated 5 jolts (difference of 1).
- Similarly, the next choices would need to be the adapter rated 6 and then the adapter rated 7 (with difference of 1 and 1).
- The only adapter that works with the 7-jolt rated adapter is the one rated 10 jolts (difference of 3).
- From 10, the choices are 11 or 12; choose 11 (difference of 1) and then 12 (difference of 1).
- After 12, only valid adapter has a rating of 15 (difference of 3), then 16 (difference of 1), then 19 (difference of 3).
- Finally, your device's built-in adapter is always 3 higher than the highest adapter, so its rating is 22 jolts (always a difference of 3).
In this example, when using every adapter, there are 7 differences of 1 jolt and 5 differences of 3 jolts.

Here is a larger example:

28
33
18
42
31
14
46
20
48
47
24
23
49
45
19
38
39
11
1
32
25
35
8
17
7
9
4
2
34
10
3

In this larger example, in a chain that uses all of the adapters, there are 22 differences of 1 jolt and 10 differences of 3 jolts.

Find a chain that uses all of your adapters to connect the charging outlet to your device's built-in adapter and count the joltage differences between the charging outlet, the adapters, and your device. 

What is the number of 1-jolt differences multiplied by the number of 3-jolt differences?

```{r get datea}
input <- as.integer(readLines("20-1210_input_adapter_array.txt"))
```


## Exercise 1

```{r}
# add in last (computer) adapter, always 3 jolts higher than highest adapter, and socket, 0, in list
x <- sort(c(0, input, max(input) + 3))

y <- table(diff(x))

# 1 volt multiplied by 2 volt differences
y["1"] * y["3"]

```

`1917`



## Exercise 2

To completely determine whether you have enough adapters, you'll need to figure out how many different ways they can be arranged. Every arrangement needs to connect the charging outlet to your device. The previous rules about when adapters can successfully connect still apply.

The first example above (the one that starts with 16, 10, 15) supports the following arrangements:

`(0), 1, 4, 5, 6, 7, 10, 11, 12, 15, 16, 19, (22)`

`(0), 1, 4, 5, 6, 7, 10, 12, 15, 16, 19, (22)`

`(0), 1, 4, 5, 7, 10, 11, 12, 15, 16, 19, (22)`

`(0), 1, 4, 5, 7, 10, 12, 15, 16, 19, (22)`

`(0), 1, 4, 6, 7, 10, 11, 12, 15, 16, 19, (22)`

`(0), 1, 4, 6, 7, 10, 12, 15, 16, 19, (22)`

`(0), 1, 4, 7, 10, 11, 12, 15, 16, 19, (22)``

`(0), 1, 4, 7, 10, 12, 15, 16, 19, (22)`

(The charging outlet and your device's built-in adapter are shown in parentheses.) Given the adapters from the first example, the total number of arrangements that connect the charging outlet to your device is 8.

The second example above (the one that starts with 28, 33, 18) has many arrangements. Here are a few:

`(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31,`
`32, 33, 34, 35, 38, 39, 42, 45, 46, 47, 48, 49, (52)`

`(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31,`
`32, 33, 34, 35, 38, 39, 42, 45, 46, 47, 49, (52)`

`(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31,`
`32, 33, 34, 35, 38, 39, 42, 45, 46, 48, 49, (52)`

`(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31,`
`32, 33, 34, 35, 38, 39, 42, 45, 46, 49, (52)`

`(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31,`
`32, 33, 34, 35, 38, 39, 42, 45, 47, 48, 49, (52)`

`(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45,`
`46, 48, 49, (52)`

`(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45,`
`46, 49, (52)`

`(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45,`
`47, 48, 49, (52)`

`(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45,`
`47, 49, (52)`

`(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45,`
`48, 49, (52)`

In total, this set of adapters can connect the charging outlet to your device in 19208 distinct arrangements.

You glance back down at your bag and try to remember why you brought so many adapters; there must be *more than a trillion* valid ways to arrange them! Surely, there must be an efficient way to count the arrangements.

What is the total number of distinct ways you can arrange the adapters to connect the charging outlet to your device?

```{r}
# req is for elements that we have to have
# first is the first and last elements (outlet and computer adapter)
req <- c(1, length(x))

# any element that has a jolt diff of 3 will require we keep both elements that constitute that difference (we can't have a jolt > 3)

# for the first in the pair for jolt 3
fp <- which(diff(x) == 3)

# add the first and second in the pair (first in pair + 1 ) to required values

req <- 
  c(req, fp, fp + 1) %>% 
  unique() %>% 
  sort()

req


# t will be our table of optional values
# that we may or may not be able to remove
# I'm going to use a for loop for each "grouping" 
#   wherein each "group" is between two other required values

t <- 
  as_tibble(x = x) %>% 
  rowid_to_column() %>% 
  mutate(req = if_else(rowid %in% req, "Y", "n"),
         group = cumsum(str_detect(req, "Y"))) %>% 
  filter(req == "n")
  
# 20 distinct groups
n_distinct(t$group)
unique(t$group)



# track results of loop
grp_results <- 
  distinct(select(t, group)) %>% 
  mutate(ncombos = 0)

for (i in unique(t$group)) {
  
  # z is our group of values we'll be looking at
  z <- t %>% filter(group == i)
  
  # int_tbl is our full list with our grouping values removed
  int_tbl <- x[-z$rowid]
  
  # does it work removing all the group values?
  if (sum(diff(int_tbl) > 3) == 0) {
    grp_results$ncombos[grp_results$group == i] <- 
      grp_results$ncombos[grp_results$group == i] + 1
  }
  
  
  # now we'll loop through each groups unique combinations
  
  for (j in seq_along(z$rowid)) {
    
    if(length(z$value) == 1) {
      combos <- list(z$value)
    } else {
      combos <- 
        combn(z$value, j, simplify = FALSE) 
    }
    
    calc <- map(combos, ~ diff(sort(c(int_tbl, .x))) )  
    
    n <- reduce(map(calc, ~ sum(.x > 3) == 0), `+`) 
    
    grp_results$ncombos[grp_results$group == i] <- 
      grp_results$ncombos[grp_results$group == i] + n
    
  }
  
}

grp_results %>% 
  # filter any that don't have any combos b/c multiplying by zero would result in zero
  filter(ncombos > 0) %>% 
  pull(ncombos) %>% 
  prod()

```

`113387824750592` unique possible combinations