---
title: '20-1214_Docking_Data'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(binaryLogic)

#disable scientific notation
options(scipen = 999)
```

## Day 13: Docking Data

As your ferry approaches the sea port, the captain asks for your help again. The computer system that runs this port isn't compatible with the docking program on the ferry, so the docking parameters aren't being correctly initialized in the docking program's memory.

After a brief inspection, you discover that the sea port's computer system uses a strange bitmask system in its initialization program. Although you don't have the correct decoder chip handy, you can emulate it in software!

The initialization program (your puzzle input) can *either update the bitmask or write a value to memory*. Values and memory addresses are both 36-bit unsigned integers. For example, ignoring bitmasks for a moment, a line like `mem[8] = 11` would write the value `11` to memory address `8`.

The bitmask is always given as a string of `36` bits, written with the most significant bit (representing `2^35`) on the left and the least significant bit (`2^0`, that is, the 1s bit) on the right. The current bitmask is applied to values immediately before they are written to memory: a `0` or `1` overwrites the corresponding bit in the value, while an X leaves the bit in the value unchanged.

For example, consider the following program:

`mask = XXXXXXXXXXXXXXXXXXXXXXXXXXXXX1XXXX0X`
`mem[8] = 11`
`mem[7] = 101`
`mem[8] = 0`

This program starts by specifying a bitmask (`mask = ....`). The mask it specifies will overwrite two bits in every written value: the `2s` bit is overwritten with `0`, and the `64s` bit is overwritten with `1`.

The program then attempts to write the value 11 to memory address 8. By expanding everything out to individual bits, the mask is applied as follows:

`value:  000000000000000000000000000000001011  (decimal 11)`
`mask:   XXXXXXXXXXXXXXXXXXXXXXXXXXXXX1XXXX0X`
`result: 000000000000000000000000000001001001  (decimal 73)`

So, because of the mask, the value `73` is written to memory address `8` instead. Then, the program tries to write `101` to address `7`:

`value:  000000000000000000000000000001100101  (decimal 101)`
`mask:   XXXXXXXXXXXXXXXXXXXXXXXXXXXXX1XXXX0X`
`result: 000000000000000000000000000001100101  (decimal 101)`

This time, the mask has no effect, as the bits it overwrote were already the values the mask tried to set. Finally, the program tries to write 0 to address 8:

`value:  000000000000000000000000000000000000  (decimal 0)`
`mask:   XXXXXXXXXXXXXXXXXXXXXXXXXXXXX1XXXX0X`
`result: 000000000000000000000000000001000000  (decimal 64)`

`64` is written to address `8` instead, overwriting the value that was there previously.

To initialize your ferry's docking program, you need the sum of all values left in memory after the initialization program completes. (The entire 36-bit address space begins initialized to the value 0 at every address.) In the above example, only two values in memory are not zero - `101` (at address `7`) and `64` (at address `8`) - producing a sum of 165.

Execute the initialization program. *What is the sum of all values left in memory after it completes?*


## INPUT

```{r input}
input_raw <- 
  read_lines("20-1214_input_docking_data.txt")

input <- 
  extract(tibble(x = input_raw), x, into = c("name", "value"), regex = "(.*) = (.*)") %>% 
  extract(col = name, into = c("type", "location"), regex = "(.*)\\[(.*)\\]", remove = FALSE) %>% 
  mutate(type = if_else(name == "mask", "mask", type), 
         location = as.numeric(location)) %>% 
  select(-name)

input <- 
  input %>% 
  mutate(mask = if_else(type == "mask", value, NA_character_)) %>% 
  fill(mask, .direction = "down") %>%
  filter(type == "mem") %>% 
  mutate(value = parse_number(value))
```


## EXERCISE 1

```{r exercise 1}

addrmax <- max(input$location[!is.na(input$location)])

bit36_init <- str_c(rep("0", 36), collapse = "")

# initialize our results to vector length of maximum memory value that will get changed, each one initialilzed to a 36 bit value set to 0
res1 <- rep(bit36_init, addrmax)

# function to convert integer to 36 bit binary number
as_binary <- function(x) {
  r <- rev(as.integer(intToBits(x)))
  pad <- 36 - length(r)
  r <- c(rep(0L, pad), r)
  return(r)
}


# need this as strtoi function has overflow error to NA
as_int_from_binary <- function(x) {
  powers <- 2 ^ rev(seq_along(x) - 1)
  sum(x * powers)
}

ex1 <- 
  input %>% 
  mutate(binary_value = map(value, as_binary), 
         mask_parse = map(mask, ~ str_split(., "")[[1]] %>% str_replace(., "X", NA_character_) %>%  as.integer()), 
         masked_value = map2(.x = mask_parse, .y = binary_value, .f = coalesce), 
         final_value = map_dbl(masked_value, as_int_from_binary)) 


for (i in seq_along(ex1$type)) {
  
  res_location <- ex1$location[i]
  
  res1[res_location] <- ex1$final_value[i]
  
}

sum(parse_number(res1))

  
```

Sum of all values in address memory returns `18630548206046`


## EXERCISE 2 NARRATIVE



## EXERCISE 2